Integer-N Frequency Synthesizer Designed for Settling Time Compliance

### 2025-05-28 CaseStudies/Example3_3_IntegerNSynthesizerDesignedforSettlingTime/article/article.qmd

@@

f0=2.4e9
f=2.4835e9
t_s = 225e-6
f_ref = 1e6
```

@@

f_range=f-f0
tuning_range_percent = 9.6 
tuning_range = f0/tuning_range_percent
V=2
Kvco = 250e6 # Hz/V
Kvco_ang = Kvco*2*math.pi
```

@@

Assuming a supply voltage of `{python} format(V)`V, a VCO gain of `{python} format(Kvco)`Hz/V (or `{python} format(Kvco_ang)`rad/s/V) is expected. The free-running frequency of the VCO is set at the minimum frequency. The center frequency $f_c$ is given by:

@@

[Eq. @eq-instant-maximum-frequency-step] represents the instant at which [Eq. @eq-inverse-laplace-phase-error] reaches its maximum value.

@@

First of all, instead of using the [Eq. @eq-inverse-laplace-phase-error] and apply $t_m$, it's calculated the maximum normalized phase error, which is the same equation normalized by the input signal ($\Delta\omega$) and the natural frequency ($\omega_n$). So the maximum normalized phase error, $\theta_{e_\text{m,n}}$, is given by:

@@

>>>>

def format_graph(value):
    abs_value = abs(value)
    if abs_value >= 1e9:
        result = f"{value / 1e9:.2f} G"
    elif abs_value >= 1e6:
        result = f"{value / 1e6:.2f} M"
    elif abs_value >= 1e3:
        result = f"{value / 1e3:.2f} k"
    elif abs_value >= 1:
        result = f"{value * 1e0:.2f} "
    elif abs_value >= 1e-3:
        result = f"{value * 1e3:.2f} m"
    elif abs_value >= 1e-6:
        result = f"{value * 1e6:.2f} μ"
    elif abs_value >= 1e-9:
        result = f"{value * 1e9:.2f} n"
    elif abs_value >= 1e-12:
        result = f"{value * 1e12:.2f} p"
    elif abs_value >= 1e-15:
        result = f"{value * 1e15:.2f} f"
    else:
        result = f"{value:.0e} "

    return result
```

@@

f_o = f0=2.4e9
f=2.4835e9
t_s = 225e-6
f_r = f_ref = 1e6
```

@@

f_range=f-f0
tuning_range_percent = 9.6 
tuning_range = f0/tuning_range_percent
Vdd=2
Kvco = 250e6 # Hz/V
Kvco_ang = Kvco*2*math.pi
```

@@

Assuming a supply voltage of `{python} format(Vdd)`V, a VCO gain of `{python} format(Kvco)`Hz/V (or `{python} format(Kvco_ang)`rad/s/V) is expected. The free-running frequency of the VCO is set at the minimum frequency. The center frequency $f_c$ is given by:

@@

[Eq. @eq-instant-maximum-frequency-step] represents the instant at which the phase error reaches its maximum value.

@@

The maximum normalized phase error, $\theta_{e_\text{m,n}}$, is given by:

@@

# Simulation

<!--```{python}
#| echo: false

import numpy as np

# Parâmetros iniciais
f_r = 10e6
f_o = 50e6
N = 5
KVCO = 10e6
I = 1e-3
Kphase = I / (2 * np.pi)
A0 = 1
K = A0 * Kphase * KVCO / N
```-->

import numpy as np

# Parâmetros iniciais
Kphase = I_CP / (2 * np.pi)
A0 = 1
K = A0 * Kphase * Kvco / N0
```

<!--```{python}
#| echo: false

from scipy import signal

omega_n = omega_bw / np.sqrt(1 + 2 * zeta**2 + np.sqrt(4 * zeta**4 + 4 * zeta**2 + 2))
f_n = omega_n/2/np.pi

# Parâmetros do PLL
C1 = K / omega_n**2
C2 = C1 / 10
R1 = 2 * zeta * omega_n / K
```-->

import math
import numpy as np
from scipy import signal

# Análise de resposta em frequência
omega_n_meas = np.sqrt(K / C_1)
f_n_meas = omega_n_meas / (2 * np.pi)

zeta_meas = (R_1 / 2) * np.sqrt(K * C_1)

t_m_meas = 1 / (omega_n_meas * math.sqrt(1 - zeta_meas**2)) * math.atan(math.sqrt(1 - zeta_meas**2) / zeta_meas)

theta_e_m_n_meas = math.sin(omega_n_meas * math.sqrt(1 - zeta_meas**2) * t_m_meas) / \
                   (math.sqrt(1 - zeta_meas**2)) * math.e**(-zeta_meas * omega_n_meas * t_m_meas)

theta_e_m_meas = 2 * math.pi

omega_bw_meas = omega_n_meas * np.sqrt(1 + 2 * zeta_meas**2 + np.sqrt(4 * zeta_meas**4 + 4 * zeta_meas**2 + 2))
f_bw_meas = omega_bw_meas / (2 * np.pi)

step_m_ang_meas = theta_e_m_meas * omega_n_meas / theta_e_m_n_meas
step_m_lin_meas = step_m_ang_meas / (2 * math.pi)

t_s_meas = math.log(0.01 * math.sqrt(1 - zeta_meas**2)) / (zeta_meas * omega_n_meas) * (-1)

# Numerador e denominador do compensador F(s)
# F(s) = (s*C1*R1 + 1) / (s^2*C1*C2*R1 + s*(C1 + C2))
F_num = [C_1 * R_1, 1]
F_den = [C_1 * C_2 * R_1, C_1 + C_2, 0]

F_sys = signal.TransferFunction(F_num, F_den)

# F(s) * K / s = G(s)
G_num = np.polymul(F_num, [K])
G_den = np.polymul(F_den, [1, 0])  # Multiplica por s

# H(s) = G(s) / (1 + G(s))
H_num = G_num
H_den = np.polyadd(G_den, G_num)

# Criando o sistema de tempo contínuo
H_sys = signal.TransferFunction(H_num, H_den)

```{python}
#| echo: false
#| label: fig-pll-step-response
#| fig-cap: "Closed-loop step response of the designed PLL."
#| fig-align: center

import numpy as np
import matplotlib.pyplot as plt
from scipy import signal

# Tempo de simulação
t = np.linspace(0, 300e-6, 100000)

# Criar vetor da entrada (degrau de amplitude 5)
step_amplitude = step_m_ang_meas
u = np.ones_like(t) * step_amplitude

# Simulação da resposta ao degrau via lsim
t_out, y_out, _ = signal.lsim(H_sys, U=u, T=t)

# Valor final da resposta
final_value = y_out[-1]
tol = 0.01 * final_value  # 1% da resposta final

# Encontrar o tempo em que a resposta entra e permanece na faixa de ±1%
ts_graphical = 0
within_bounds = np.abs(y_out - u) <= tol
for i in range(len(within_bounds)):
    if all(within_bounds[i:]):
        ts_graphical = t_out[i]
        break

# Gerar gráfico
plt.figure(figsize=(8, 5))
plt.plot(t_out * 1e6, y_out, label='Step Response')
plt.plot(t_out * 1e6, u, linestyle=':', color='black', label='Step')
plt.axhline(final_value + tol, color='red', linestyle='--', linewidth=0.8, label='±1% Tolerance')
plt.axhline(final_value - tol, color='red', linestyle='--', linewidth=0.8)
plt.axvline(ts_graphical * 1e6, color='green', linestyle='--', linewidth=1.2, label=f'Graphical settling time: {ts_graphical*1e6:.5f} us')
plt.xlabel('Time [us]')
plt.ylabel('Phase [rad]')
plt.title('Step Response with Settling Time (99%)')
plt.grid(True)
plt.legend()
plt.tight_layout()
plt.show()
```

Using the parameters $\omega_n$ and $\zeta$, the transfer function of the entire loop can be described as (neglecting the second capacitor, $C_2$, of the filter):

$$
\frac{\theta_o}{\theta_R}=\frac{\omega_n^2\left(\frac{2\zeta}{\omega_n}s+1\right)}{s^2+2\zeta\omega_n s+\omega_n^2}
$$ {#eq-full-transfer-function-no-C2}

The natural frequency of the loop, $\omega_n$, was computed from the loop bandwidth $\omega_\mathrm{bw}$ and damping factor $\zeta$, yielding a natural frequency $f_n=\omega_n/2\pi=$ `{python} format(f_n)`Hz.

Based on the chosen parameters, the loop filter components were calculated as follows. The primary capacitor of the loop filter is:

$$
C_1 = \frac{K_\theta K_\mathrm{VCO}}{N \omega_n^2}
$$

That is, $C_1=$ `{python} format(C_1)`F.

The second capacitor is chosen to be one-tenth of $C_1$, i.e., $C_2=$ `{python} format(C_2)`F.

The resistor of the loop filter is given by:

$$
R_1 = \frac{\zeta 4 \pi N \omega_n}{I K_\mathrm{VCO}}
$$

That is, $R_1=$ `{python} format(R_1)`$\Omega$.

These component values define the loop filter required to achieve the desired dynamic behavior of the PLL. With these components, it's possible to extract another set of the system parameters, i.e., $f_n=\omega_n/2\pi$, $\zeta$ and $f_\mathrm{bw}=\omega_\mathrm{bw}/2\pi$. The values are $f_n=$ `{python} format(f_n_meas)`Hz, $\zeta=$ `{python} round(zeta_meas, 3)` and $f_\mathrm{bw}=$ `{python} format(f_bw_meas)`Hz.

Similarlly what was presented in [Eq. @eq-full-transfer-function-no-C2], it's possible to present the transfer function of the system in terms of components values. This transfer function can be seen below (this equation uses the second capacitor, $C_2$, of the loop filter, adding a high-frequency pole to the system):

$$
\begin{split}
\frac{\theta_o}{\theta_R}&=\frac{K_\mathrm{VCO}K_\theta\left(1+sC_1R\right)}{s^2N\left(C_1+C_2\right)\left(1+sC_sR\right)+K_\mathrm{VCO}K_\theta\left(1+sC_1R\right)}
\end{split}
$$

where $C_s$ is the paralel between $C_1$ and $C_2$. With the PLL values properly defined, the transfer function of this system is given by:

from scipy import signal
from IPython.display import display, Markdown

def format_eng(x):
    if x == 0:
        return "0"
    exponent = int("{:.0e}".format(x).split('e')[1])
    base = x / (10**exponent)
    return f"{base:.2f}e{exponent}"

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)

# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = H_sys.num[0][0] if isinstance(H_sys.num[0], (list, tuple)) else H_sys.num
den = H_sys.den[0][0] if isinstance(H_sys.den[0], (list, tuple)) else H_sys.den

num_str = poly_to_str(num)
den_str = poly_to_str(den)

# Exibe como equação no Jupyter
display(Markdown(f"$$H(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-transfer-function}}"))

from scipy import signal
from IPython.display import display, Markdown

def format_eng(x):
    if x == 0:
        return "0"
    exponent = int("{:.0e}".format(x).split('e')[1])
    base = x / (10**exponent)
    return f"{base:.2f}e{exponent}"

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)

# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = F_sys.num[0][0] if isinstance(F_sys.num[0], (list, tuple)) else F_sys.num
den = F_sys.den[0][0] if isinstance(F_sys.den[0], (list, tuple)) else F_sys.den

num_str = poly_to_str(num)
den_str = poly_to_str(den)

# Exibe como equação no Jupyter
display(Markdown(f"$$F(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-transfer-function-loop-filter}}"))

A step with the same amplitude as the previously calculated frequency step was also applied in phase, that is, $\Delta\theta=$ `{python} format_with_pi_latex(step_m_ang_meas)`rad/s. Applying this signal to [Eq. @eq-transfer-function], the response of this system to the phase step is obtained. With this step, the settling time achieved is $t_s=$ `{python} format(t_s_meas)`s. Through simulation, it is possible to verify that the obtained value is $t_s=$ `{python} format(ts_graphical)`s. The simulated response can be visualized in [Fig. @fig-pll-step-response].

import numpy as np
import matplotlib.pyplot as plt
from scipy import signal

# Gerar a resposta em frequência
w, mag, phase = signal.bode(H_sys)

# Encontrar o ponto onde a magnitude cruza -3 dB
cutoff_index = np.argmax(mag <= -3)
cutoff_freq_ang = w[cutoff_index]
cutoff_freq_lin = cutoff_freq_ang/2/np.pi
cutoff_mag = mag[cutoff_index]

# Plotando o diagrama de Bode
plt.figure(figsize=(10, 6))

# Magnitude
plt.subplot(2, 1, 1)
plt.semilogx(w, mag, label='Magnitude (dB)')
plt.axvline(x=cutoff_freq_ang, color='r', linestyle='--', label='Cutoff Frequency (-3 dB)')
plt.axhline(y=-3, color='gray', linestyle=':', label='-3 dB')
plt.scatter(cutoff_freq_ang, cutoff_mag, color='red')  # Marcar o ponto

# Adiciona o valor da frequência no gráfico
plt.text(cutoff_freq_ang, cutoff_mag - 5, f'{format_graph(cutoff_freq_lin)}Hz', color='red',
         ha='center', va='top', fontsize=9, bbox=dict(facecolor='white', edgecolor='red', boxstyle='round'))

plt.title('Frequency Response - Bode Plot')
plt.ylabel('Magnitude (dB)')
plt.grid(True, which="both", ls="--")
plt.legend()

# Fase
plt.subplot(2, 1, 2)
plt.semilogx(w, phase)
plt.axvline(x=cutoff_freq_ang, color='r', linestyle='--')
plt.ylabel('Phase (degrees)')
plt.xlabel('Frequency (rad/s)')
plt.grid(True, which="both", ls="--")

plt.tight_layout()
plt.show()
```

import numpy as np
import matplotlib.pyplot as plt
from scipy import signal

# Gerar a resposta em frequência
w, mag, phase = signal.bode(F_sys)

# Encontrar o ponto onde a magnitude cruza -3 dB
cutoff_index = np.argmax(mag <= -3)
cutoff_freq_ang = w[cutoff_index]
cutoff_freq_lin = cutoff_freq_ang/2/np.pi
cutoff_mag = mag[cutoff_index]

# Plotando o diagrama de Bode
plt.figure(figsize=(10, 6))

# Magnitude
plt.subplot(2, 1, 1)
plt.semilogx(w, mag, label='Magnitude (dB)')

plt.title('Bode Plot - Loop filter')
plt.ylabel('Magnitude (dB)')
plt.grid(True, which="both", ls="--")
plt.legend()

# Fase
plt.subplot(2, 1, 2)
plt.semilogx(w, phase)
plt.ylabel('Phase (degrees)')
plt.xlabel('Frequency (rad/s)')
plt.grid(True, which="both", ls="--")

plt.tight_layout()
plt.show()
```

## Cycle slipping analyzing

import numpy as np

omega_r_safe = omega_n*10
f_r_safe = omega_r_safe/2/np.pi

omega_r_max = omega_n * 2 * np.pi * zeta
f_r_max = omega_r_max/2/np.pi
```

With the intention of conducting tests to verify the occurrence of cycle slipping, the minimum frequency of the reference signal is estimated. A ratio often quoted as being safe is 10:1 for $\omega_r/\omega_n$; thus, the minimum safe frequency for the reference signal is given by $f_r=$ `{python} format(f_r_safe)`Hz. The minimum allowed frequency is given by:

$$
\frac{\omega_r}{\omega_n}\ge 2\pi\zeta
$$

That is, the minimum reference frequency is $f_r\ge$ `{python} format(f_r_max)`Hz. Therefore, since a PLL operating with a reference frequency of $f_r=$ `{python} format(f_r)`Hz and an output frequency of `{python} format(f_o)`Hz is desired, the designed system must operate correctly.

Using the Simulink software, a testbench was prepared to verify the settling time of the PLL. With the previously calculated parameters and using the maximum step frequency value, the resulting curve can be seen in [Fig. @fig-max-step-response]. It can be observed that the step is applied at the instant of 1 ms, as the system is already properly stabilized at that moment. As predicted, cycle slipping does not occur; however, unlike what was calculated, the settling time of this signal is $9.196\;\mu\text{s}$.

<!--![System response to the maximum step frequency](../images/max_step_frequency_response.jpg){#fig-max-step-response}-->

cycle_slip_step_lin = 3*step_m_lin
```

Moreover, contrary to what was previously predicted, for cycle slipping to occur, the step frequency must be approximately $\Delta f=$ `{python} format(cycle_slip_step_lin)`Hz. The system's response to this signal can be seen in [Fig. @fig-cycle-slip-response].

<!--![System response with cycle slip](../images/step_frequency_response_cycle_slip.jpg){#fig-cycle-slip-response}-->
